Fast Sorting on Modern Computers

### 2005-11-01 docs/tech_report/conclusions.tex
did the abstract. removed comment from the latex files
uncommitted changes from ages ago

In several cases, we found that direct-mapped caches performed better than
fully-associative ones, resulting in fewer misses. This was due to they're
replacement policy, and that the sorts in question were heavily optimized for
the direct-mapped caches. None of the changes slowed the fully-associative
caches however.

@@

\afterpage{
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-10em}
%\setcounter{figure}{9} % note the cheating to make this work
\begin{figure}[H]
\begin{changemargin}
\includegraphics[angle=90,scale=1.8]{plots/all_cycles}
\vspace*{8em}
\end{changemargin}
\caption[Cycles per key of several major sorts and their variations]{\label{All
cycles}Cycles per key of several major sorts and their variations - this was
measured on a Pentium 4 using hardware performance counters.}
\end{figure}
\newpage
}

The best stable sort is memory-tuned radixsort. The best comparison-based stable
sort is double-aligned tiled mergesort. The best in-place sort is base
quicksort. The best in-place stable sort we examine in this report is insertion
sort. The best elementary sort is also insertion sort; as such it should be only
used for very small data sets. The best sort for very large data sets is
memory-tuned radixsort. The fastest overall sort was memory-tuned radixsort,
though all the base and memory-tuned quicksort are both very close.

>>>>

In several cases, we found that direct-mapped caches performed better than
fully-associative ones, resulting in fewer misses. This was due to their
replacement policy, and that the sorts in question were heavily optimized for
the direct-mapped caches. None of the changes slowed the fully-associative
caches however.

@@

\afterpage{
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-10em}
\begin{figure}[H]
\begin{changemargin}
\includegraphics[angle=90,scale=1.8]{plots/all_cycles}
\vspace*{8em}
\end{changemargin}
\caption[Cycles per key of several major sorts and their variations]{\label{All
cycles}Cycles per key of several major sorts and their variations - this was
measured on a Pentium 4 using hardware performance counters.}
\end{figure}
\newpage
}

The best stable sort is memory-tuned radixsort. The best comparison-based stable
sort is double-aligned tiled mergesort. The best in-place sort is base
quicksort. The best in-place stable sort we examine in this report is insertion
sort. The best elementary sort is also insertion sort; as such it should be only
used for very small data sets. The best sort for very large data sets is
memory-tuned radixsort. The fastest overall sort was memory-tuned radixsort,
though the base and memory-tuned quicksorts are all very close.

### 2005-11-01 docs/tech_report/intro.tex
widenned the pages
uncommitted changes from ages ago

Papers considering sorting performance typically analyse instruction count.
LaMarca and Ladner \cite{LaMarca96} break this trend by discussing the cache
implications of several algorithms, including sorts. This report investigates
these claims and tests and analyses high performance sorts based upon their
work.

@@

Chapters \ref{heap} through \ref{radix} discuss base algorithms and
cache-conscious improvements for heapsort, mergesort and quicksort. Results are
presented and explained, and branch prediction implications of the changes are
discussed.

@@

Chapter \ref{conclusions} presents conclusions, summarises results from previous
chapters, compares our results with LaMarcas, presents a list of cache and
branch prediction results and improvements, discusses the fastest sort in our
tests and lists contributions made by this research.

>>>>

Papers considering sorting performance typically analyse instruction count.
LaMarca and Ladner break this trend in \cite{LaMarca96} by discussing the cache
implications of several algorithms, including sorts. This report investigates
these claims and tests and analyses high performance sorts based upon their
work.

@@

Chapters \ref{heap} through \ref{radix} discuss base algorithms and
cache-conscious improvements for heapsort, mergesort, quicksort and radixsort.
Results are presented and explained, and branch prediction implications of the
changes are discussed.

@@

Chapter \ref{conclusions} presents conclusions, summarises results from previous
chapters, compares our results with LaMarca's, presents a list of cache and
branch prediction results and improvements, discusses the fastest sort in our
tests and lists contributions made by this research.

### 2005-11-01 docs/tech_report/merge.tex
lots of minor changes
did the abstract. removed comment from the latex files
uncommitted changes from ages ago

%\section{Description}
\label{merge}
Merging is a technique which takes two sorted lists, and combines them into a
single sorted list. Starting at the smallest key on both lists, it writes the
smaller key to a third list. It then compares the next smallest key on that list
with the next smallest on the other, and writes the smaller to the auxiliary
list. The merging continues until both lists are combined into a single sorted
list.

@@

\label{mergesort replacement policy}
It is notable that our prediction of half a miss per key for multi-mergesort,
which was taken from LaMarca, is exactly right. However, our assertion that the
fully-associative cache will perform slightly better is false. In actual fact,
the direct-mapped cache performs significantly better (see Figure
\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
our fully-associative cache removes the least recently used keys from the cache
to make room, while the direct-mapped caches are designed to make best use of
the keys while they are in the cache. They have, effectively, an optimised
replacement policy.

@@

Figure \ref{mergesorts-1} shows the behaviour of the individual branches within
algorithm N. The labels in the graphs correspond to the names of the steps in
Knuth's algorithm (we also used the same names for the labels of branch targets
in our C source code. Branch N3 is the main comparison branch; it is used to
choose the smaller of the elements from the start of the two sub-arrays being
merged. The direction of this branch is completely random - it depends entirely
on the (in this case random) data. Figure \ref{mergesorts-1} also shows that
this branch is almost entirely unpredictable using a bimodal predictor. The
result for a two-level predictor is much the same, because there is no pattern
in the outcome of these branches. Branches `\n{N3 i==j}' (check for the current
merge being complete), `\n{N5}' and `\n{N9} (both checks for the end of a run of
increasing elements) are control-flow branches, and are almost entirely
predictable. Figure \ref{mergesorts-2} shows that the comparison branch in
Algorithm S is just as unpredictable as that in algorithm N. However, given that
the merge in algorithm S is entirely regular (see section \ref{Algorithm S}) it
is possible to eliminate one of the control-flow branches.

>>>>

\label{merge}
Merging is a technique which takes two sorted lists, and combines them into a
single sorted list. Starting at the smallest key on both lists, it writes the
smaller key to a third list. It then compares the next smallest key on that list
with the next smallest on the other, and writes the smaller to the auxiliary
list. The merging continues until both lists are combined into a single sorted
list.

@@

\label{mergesort replacement policy}
It is notable that our prediction of half a miss per key for multi-mergesort,
which was taken from LaMarca, is exactly right. However, our assertion that the
fully-associative cache will perform slightly better is false. In actual fact,
the direct-mapped cache performs significantly better (see Figure
\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
our fully-associative cache removes the least recently used keys from the cache
to make room, while the direct-mapped caches are designed to make best use of
the keys while they are in the cache. They have, effectively, an optimized
replacement policy.

@@

Figure \ref{mergesorts-1} shows the behaviour of the individual branches within
algorithm N. The labels in the graphs of algorithms N and S correspond to the
names of the steps in Knuth's algorithm (we also used the same names for the
labels of branch targets in our C source code. Branch `\n{N3}' is the main
comparison branch; it is used to choose the smaller of the elements from the
start of the two sub-arrays being merged. The direction of this branch is
completely random - it depends entirely on the (in this case random) data.
Figure \ref{mergesorts-1} also shows that this branch is almost entirely
unpredictable using a bimodal predictor. The result for a two-level predictor is
much the same, because there is no pattern in the outcome of these branches.
Branches `\n{N3 i==j}' (check for the current merge being complete), `\n{N5}'
and `\n{N9} (both checks for the end of a run of increasing elements) are
control-flow branches, and are almost entirely predictable. Figure
\ref{mergesorts-2} shows that the comparison branch in Algorithm S is just as
unpredictable as that in algorithm N. However, given that the merge in algorithm
S is entirely regular (see section \ref{Algorithm S}) it is possible to
eliminate one of the control-flow branches.
